Fusing Filters with Integer Linear Programming

### 2014-05-13 2014betterfusionforfilters/section/01-Introduction.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?
fill in some fusion refs in intro

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}
A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.  It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance. This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{some fusion refs}, to name just a few. Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures. As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering. Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.  The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.

@@

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
The function @normalize@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.

>>>>

%!TEX root = ../Main.tex
\section{Introduction}
\label{s:Introduction}
A collection-oriented programming model is well suited for expressing data-parallel computations, as it exposes the communication patterns to the compiler without the need for complicated loop analysis.  It is essential, however, that the compiler combines those operations and compiles them into a small number of efficient imperative loops as a naive, operation-by-operation translation, would lead to high memory consumption and memory traffic, and result in poor performance. This is a well known problem, and techniques to fuse subsequent array operations into a single, more complex one, have been presented in \cite{coutts2007streamfusion, gao1993collective, kennedy1993typedfusion, kennedy2001fastgreedy, keller2010repa}, to name just a few. Fusion is, in general, not sufficient to minimise memory traffic: first, we often need to re-arrange the array operations and cluster those that iterate over the same structures. As shown in  (\cite{megiddo1998optimal} and Darte~\cite{darte2002contraction}), Integer Linear Programming can be used to find such a clustering. Unfortunately, since they cannot handle operations like filter, which produce arrays whose size differs from the input arrays.  The technique we present in this paper can handle both multi-loop fragments as well as size-altering operations.

@@

To demonstrate the effect of clustering and fusion, consider the following code snippet:
\begin{figure}[H]
\begin{code}
 normalize2 :: Array Int -> Array Int
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ys1  = map    (/ sum1) xs
        ys2  = map    (/ sum2) xs
    in (ys1, ys2)
\end{code}
\end{figure}
The function @normalize2@ computes two sums, one of all the elements of @xs@, the other of all the ones greater than zero. Since we need to fully evalute the sums to proceed with the maps, it is clear that we need at least two separate loops. These folds are examples of fusion-preventing dependencies, as fold cannot be fused with subsequent operations. Figure~\ref{f:normalize2-clusterings} shows the data-flow graph of @normalize@. In the leftmost diagram we can see the effect of applying stream fusion to the program: we end up with four loops (denoted by the dotted lines): only the filter operation is combined with the subsequent fold. The best existing ILP approach results in the right-most graph: it combines the @sum1@ fold and @sum2@ filter in one loop, but requires an extra loop for the fold operation which consumes the filter output, since it cannot fuse filter operations. Our approach (diagram in the middle) produces the, for this example, optimal solution: two loops. The first to calculate both sums, the second to calculate the maps.
